
\chapter{Introduction} % Main chapter title

\label{introduction} % Change X to a consecutive number; for referencing this chapter elsewhere, use \ref{ChapterX}

%----------------------------------------------------------------------------------------

% Define some commands to keep the formatting separated from the content 
\newcommand{\keyword}[1]{\textbf{#1}}
\newcommand{\tabhead}[1]{\textbf{#1}}
%\newcommand{\code}[1]{\colorbox{lightgray}{\texttt{#1}}}
\newcommand{\code}[1]{{\sethlcolor{lightgray}\hl{\texttt{#1}}}}
\newcommand{\file}[1]{\texttt{\bfseries#1}}
\newcommand{\option}[1]{\texttt{\itshape#1}}

%----------------------------------------------------------------------------------------


\section{Metagenomics}

Since the human genome project, where in a massive collaborative effort the first human genome has been sequenced and assembled, Next generation sequencing (NGS) methods have taken over. 
NGS methods like sequencing by synthesis, which is used in Illumina sequencing machines, enabled the creation of huge amounts of data. 
% wgs, shotgun vs 16s

The general importance of metagenomics has just become clear in the recent years. 
% examples needed
Certain illnesses are not only determined by genome and lifestyle but also by the microbiome composition of affected body parts. % which ones? 

For example some species and strains are implicated in causing inflammatory bowel disease (IBD) % more specific



\section{Machine Learning and Reproducibility}

Recent advances in computing power enabled the extensive use of Machine Learning (ML) algorithms. 
Frameworks like \keyword{scikit-learn} have become better and popular \citep{scikit}.

Algorithms like Support Vector Machines (SVM), Random Forest (RF) and Gradient-Boosted Trees (XG-Boost) have been successfully used to predict illnesses based on metagenomic data before. These studies lack a focus on reproducibility and do not have best practices in mind.

\citep{xgboost}
% FOSS important for reproducibility

\section{Thesis Objective}
Exploration of publicly available datasets 

Create a pipeline which makes conducting Machine Learning experiments on metagenomic data for phenotype prediction reproducible. The pipeline should be designed following best practices so potential users do not have to focus on this.
% taxonomic profiler
